Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2018-01-31 manuscript/response_to_reviewers.tex

\begin{enumerate}

\item{\textcolor{blue}{In the section on adding in reconstructed ancestors to the dataset before ordination, I suppose you should mention that this is similar to the approach that we outlined in our Paleobiology paper (Brusatte et al. 2011) and mention any key differences.}}

We now explicitly refer to Brusatte et al. 2011 and mention the key difference:

\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines @@@

@@

\item{\textcolor{blue}{I’m not sure about the use of the term ‘cladisto-spaces’. To me this implies that the space itself is ordinated using phylogenetic information (e.g., branch lengths). Now, you are indeed including some phylogenetic information in your morphospaces, in terms of reconstructed ancestors and inferred states for branches. So sure, I think what you have could be called a ‘cladisto-space’. Same with most of the morphospace options produced by Claddis. But the ones that most of us have done in the past, simply using discrete characters to ordinate a morphospace, haven’t generally incorporated phylogeny. I’m talking about the references you cite in lines 141-143 here. These morphospaces been ordinated using discrete characters of the type that are used in phylogenetic analysis. But the characters are not phylogenetically transformed in any way. As far as the ordination analysis knows, they are simply discrete character statements. So I wonder if the term ‘cladisto-space’ is confusing, and is just yet another new term that may muddy the waters rather than clarify. If you want to keep using this term, that’s fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}

We've change the term cladisto-space to morphospace throughout the manuscript.
Although we prefer the sound and distinction from the ``cladisto-space'' term, we agree with this reviewer concern about adding a``new term that may muddy the waters rather than clarify''.
We've emphasised the differences, however, with other uses of the word ``morphospace'' in an explanatory paragraph:

\textit{To explore disparity-through-time in our datasets, we used a morphospace approach (CITATIONS).
Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f. \textit{geometric morphometric data) and includes some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.} lines @@@

%TG: what do you think?

### 2018-01-31 manuscript/time-slice.tex
changed cladisto-space to morphospace (sigh...)

\paragraph{Building cladisto-spaces.} 
To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on discrete morphological data (i.e. the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
%TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well. Second, during the PhD, we used cladisto-space to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space... However, I like the sound of cladisto-space and most people will understand it the way we do. I'm just thinking about pedantic-palaeo-reviewers...
% NC: Good point. Let's leave it for now as I think this bit is nicely explained, but easy enough to edit if people don't like it.
Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data.

% To explore disparity-through-time in our datasets we used a discrete morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
% This approach is a approximation of a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010} and is based on discrete morphological data (i.e. the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
% Although empirically inter-taxon distances are the same in a morphospace based on discrete or continuous data \citep{foth2012different,hetherington2015cladistic}, we like to highlight that this space is based on discrete morphological data and phylogeny (it could be referred to as a ``cladisto-space'').
% Mathematically, a cladisto-space is an $n$ dimensional object that summarises the distances between the taxa present in a discrete morphological matrix.

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For discrete morphological matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

\paragraph{Ordination.}
After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}; but distinguished in \citealt{legendre2012numerical}) is an eigen decomposition of the distance matrix.
Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}). 
We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the cladisto-space) where $n$ is equal to $k-2$, i.e. the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g. selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).

Note that our cladisto-spaces represent an ordination of all possible morphologies coded in each study through time.
It is unlikely that all morphologies will co-occur at each time point, therefore, the disparity of the whole cladisto-space is expected to be greater than the disparity at any specific point in time.

@@

\paragraph{Calculating disparity.}
\label{disparity_calc}
Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however a majority of studies in palaeobiology estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
However, these metrics have limitations. 
First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves. 
We therefore use the sum of variances metric to estimate disparity here:

@@

\paragraph{Time-slicing.} 
\label{time_slicing}
The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling of the cladisto-space across time and allows us to use different underlying models of character evolution (punctuated or gradual).

@@

It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for ancestral character estimation in the ``Preparing the data for disparity-through-time analysis'' section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the methods described above.
This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the cladisto-space.
In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g. distance between species, variance of each axis, etc.) to remain constant.

>>>>

\paragraph{Building morphospaces.} 
% To explore disparity-through-time in our datasets we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
% This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a morphospace is an approximation of the morphospace based on discrete morphological data (i.e. the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
% %TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well. Second, during the PhD, we used morphospace to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space... However, I like the sound of morphospace and most people will understand it the way we do. I'm just thinking about pedantic-palaeo-reviewers...
% % NC: Good point. Let's leave it for now as I think this bit is nicely explained, but easy enough to edit if people don't like it.
% Mathematically, a morphospace is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
% Although empirically inter-taxon distances are the same in a morphospace or a morphospace \citep{foth2012different,hetherington2015cladistic}, we prefer the term morphospace to make it clear that this space is estimated using cladistic data and not morphometric data.

To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.

\paragraph{Constructing distance matrices.}
To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For discrete morphological matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

\paragraph{Ordination.}
After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}; but distinguished in \citealt{legendre2012numerical}) is an eigen decomposition of the distance matrix.
Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}). 
We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the morphospace) where $n$ is equal to $k-2$, i.e. the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our morphospaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g. selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).

Note that our morphospaces represent an ordination of all possible morphologies coded in each study through time.
It is unlikely that all morphologies will co-occur at each time point, therefore, the disparity of the whole morphospace is expected to be greater than the disparity at any specific point in time.

@@

\paragraph{Calculating disparity.}
\label{disparity_calc}
Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however a majority of studies in palaeobiology estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the morphospace \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
However, these metrics have limitations. 
First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves. 
We therefore use the sum of variances metric to estimate disparity here:

@@

\paragraph{Time-slicing.} 
\label{time_slicing}
The ``time-slicing'' approach considers subsets of taxa in the morphospace at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling of the morphospace across time and allows us to use different underlying models of character evolution (punctuated or gradual).

@@

It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for ancestral character estimation in the ``Preparing the data for disparity-through-time analysis'' section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the methods described above.
This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the morphospace.
In other words, this method does not require a re-ordination of the morphospace every time a slice goes through a branch, thus allowing the properties of the morphospace (e.g. distance between species, variance of each axis, etc.) to remain constant.
